\documentclass[11pt,a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{doi}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{array}
\usepackage{enumitem}

\geometry{margin=2.5cm}

\hypersetup{
  colorlinks=true,
  linkcolor=black,
  citecolor=black,
  urlcolor=black
}

\theoremstyle{definition}
\newtheorem{remark}{Remark}

\newcommand{\Heis}{\mathrm{Heis}}
\newcommand{\SU}{\mathrm{SU}}
\newcommand{\su}{\mathfrak{su}}
\newcommand{\SPI}{S_\Pi}
\newcommand{\cchi}{c_\chi}
\newcommand{\Vr}{V_\rho}

\title{The Non-Injective Foundations Sub-Programme\\[0.4em]
\large Presentation Note 5}

\author{J\'{e}r\^{o}me Beau\\[0.3em]
\small Independent Researcher\\
\small \texttt{jerome.beau@cosmochrony.org}}

\date{2026}

\begin{document}

  \maketitle

  \begin{abstract}
    The non-injective foundations sub-programme establishes the axiomatic primitive of the
    Cosmochrony corpus and answers a single central question: which structures are forced
    once admissible non-injective projection is taken as primitive?
    Starting from four axioms (A1--A4) governing admissible transitions between observable
    states, the sub-programme derives as theorems --- not postulates --- the following:
    the non-injectivity of $\Pi$ as a structural necessity of genuine emergence (ENI);
    the discrete Heisenberg group $\Heis_3(\mathbb{Z}/q\mathbb{Z})$ and its Weil
    representation $\Vr$ as the unique admissible fibre structure (Foundation~M,
    HeisenbergStructure);
    and the absence of any independent dimensional parameter beyond $\cchi$ (noscale).
    The sub-programme also provides the definitive liquidation of the preliminary vocabulary
    of the white paper: $\chi$, relaxation, and iterated projection dynamics are not
    postulates of the axiomatic formulation but emergent consequences of A1--A4.
  \end{abstract}

  \medskip
  \noindent\textbf{Keywords.}
  non-injective projection; axiomatic primitive; Heisenberg group; Weil representation;
  admissibility; Born--Infeld; projection entropy; no-go theorem; emergence; foundations;
  presentation note.

  \newpage
  \tableofcontents

  \newpage

  \section{Motivation and Central Question}
    \label{sec:motivation}

    The non-injective foundations sub-programme answers the following question.

    \medskip
    \noindent\textit{Central question.}
    Is it possible to derive $\Heis_3(\mathbb{Z}/q\mathbb{Z})$, its Weil representation,
    the Born--Infeld saturation constraint, and the arrow of time as consequences of a
    minimal set of structural axioms?
    More precisely:
    \textit{which structures are forced once admissible non-injective projection is taken
    as the sole primitive?}

    \medskip
    The sub-programme answers: all of them.
    Non-injectivity is not an interpretive option or a modelling choice.
    It is a structural necessity of any genuinely emergent description (ENI).
    The Heisenberg group structure is not introduced as a postulate; it is the unique
    algebraic form compatible with the joint constraint of irreducibility (A3) and
    projection locking (A4).
    The BI saturation constraint is the unique bounded completion of the admissible
    effective action.
    The arrow of time follows from A1+A2 alone.
    No background spacetime, no Hilbert space, and no quantum postulate are introduced
    at this level.

    \medskip
    This note is the definitive entry point to the corpus for a reader who wants to understand
    \emph{why} the specific structures of the other sub-programmes (spectral admissibility,
    geometry, gauge, gravity) appear, rather than merely how they are used.

  \section{Position in the Cosmochrony Programme}
    \label{sec:position}

    The non-injective foundations sub-programme is the axiomatic spine of the corpus.
    It occupies Level~1 of the three-level architecture:
    Level~1 (primitive), Level~2 (spectral fibre), Level~3 (physical observables).

    Every other sub-programme takes one or more outputs of the foundations as inputs.
    \begin{itemize}
      \item Note~1 (Spectral Admissibility) takes $\Heis_3(\mathbb{Z}/q\mathbb{Z})$ and $\Vr$,
      the BI bound $A_n \leq \cchi/\sqrt{\lambda_n}$, and the parity involution as given.
      \item Note~2 (Emergent Geometry) takes $\Vr$ and the four-dimensional Carnot convergence
      as the starting point of the Mosco limit.
      \item Note~3 (Gauge Structure) takes the phase fibre and the non-injectivity of $\Pi$ as
      the origin of gauge freedom.
      \item Note~4 (Spectral Gravity) takes $\SPI > 0$ as the structural condition that forces
      an effective gravitational sector.
    \end{itemize}

    None of these sub-programmes can be understood without the foundations.
    The foundations sub-programme provides the \emph{reason} for the structures the other
    sub-programmes compute.

  \section{From the Preliminary Vocabulary to the Axiomatic Primitive}
    \label{sec:vocab}

    \begin{remark}[Vocabulary note: liquidation of the preliminary $\chi$ framework]
      The white paper~\cite{Cosmochrony} introduced a substrate $\chi$, a relaxation
      mechanism, and an iterated projection cascade $U_{t+1} = \Pi \circ \sigma(U_t)$ as
      the engine of dynamics.
      This vocabulary is superseded by the axiomatic formulation.

      The axiomatic formulation retains the structural content while removing the dynamical
      reading:
      the primitive is not a substrate process, but the local structure of admissible
      non-injective transitions.
      The substrate $\chi$ is static; no relaxation mechanism is postulated at the
      foundational level.
      Observable evolution is an induced ordering of admissible projections, not a
      dynamics of $\chi$ itself.
      Relaxation and iteration appear as structural consequences of A1--A4, not as
      independently postulated processes.

      This liquidation is final: the term ``relaxation'' should not appear in any description of
      the axiomatic framework, and $\chi$ should not be described as dynamical.
      In all notes of this series (Notes~1--9), the admissibility constraint replaces the
      earlier relaxation vocabulary.
    \end{remark}

  \section{Logical Chain of the Sub-Programme}
    \label{sec:chain}

    The sub-programme is organised around the following derivation chain:
    \[
      \text{A1--A4}
      \;\Longrightarrow\;
      \underbrace{\Pi \text{ non-injective},\; \SPI > 0}_{\text{ENI, Theorem~3}}
      \;\Longrightarrow\;
      \underbrace{[X,Y] \neq 0}_{\text{Foundation, Thm~5.4}}
      \;\Longrightarrow\;
      \underbrace{\Heis_3(\mathbb{Z}/q\mathbb{Z})}_{\text{Foundation, Thm~5.7}}
      \;\Longrightarrow\;
      \underbrace{\Vr}_{\text{HeisStr, Thm~4.1}}.
    \]

    Four structural implications.

    \begin{enumerate}
      \item \textit{Non-injectivity from genuine emergence (ENI~\cite{Beau2026l}).}
      Given surjective $\Pi : \Omega \to O$ with $O$ not structurally isomorphic to $\Omega$
      and no microstate-resolving structure beyond $\Pi$: $\Pi$ is necessarily non-injective
      and $\SPI > 0$.
      This is a framework-independent no-go theorem, proved from surjectivity, informational
      completeness, and non-isomorphism alone.

      \item \textit{Non-trivial commutator from A3 (Foundation~\cite{Beau2026m}).}
      A3 (admissibility as non-premature selection) forces the irreducibility of the admissible
      fibre $F_n$.
      Any decomposition $F_n = F^{(1)} \oplus F^{(2)}$ stable under admissible generators is
      forbidden prior to projection locking.
      Irreducibility forces a non-trivial commutator $[X, \sigma(X)] = Z \neq 0$ between
      conjugate admissible generators $X$ and $\sigma(X)$.

      \item \textit{Heisenberg structure forced by A1--A3 and BI parity.}
      The non-trivial commutator $[X, Y] = Z$ generates a three-dimensional nilpotent Lie
      algebra.
      The centre is generated by $Z$; its prime order $q$ is the spectral resolution parameter.
      By the Stone--von Neumann uniqueness theorem, the unique irreducible unitary
      representation of $\Heis_3(\mathbb{Z}/q\mathbb{Z})$ with central character $e^{2\pi i/q}$
      is the Weil representation $\Vr$.

      \item \textit{BI parity involution.}
      Born--Infeld parity (BornInfeld~\cite{Beau2026c}) forces the minimal fibre condition
      $\Pi^{-1}(y) \supseteq \{\chi, -\chi\}$.
      Combined with A3, this closes the identification of $\Vr$ as the unique admissible fibre.
    \end{enumerate}

  \section{Constituent Papers}
    \label{sec:papers}

    \subsection{Non-injectivity as structural necessity: ENI}
      \label{sec:eni}

      \textbf{ENI}~\cite{Beau2026l} (Non-Injectivity as a Structural Necessity of Genuine
  Emergence) is the foundational no-go theorem.

      \textit{Main theorem} (Theorem~3; proved).
      Any surjective $\Pi : \Omega \to O$ that is informationally complete for observables
      and such that $O$ is not structurally isomorphic to $\Omega$ must be non-injective with
      $\SPI > 0$.
      The proof uses only surjectivity, informational completeness, and the non-isomorphism
      condition; it is independent of any physical framework.

      \textit{Categorical and informational reformulations} (§8).
      Genuine emergence corresponds to a non-faithful functor.
      The Kullback--Leibler divergence of any state distribution with respect to its projection
      lift is strictly positive.

      \textit{Corollary~6: Recursive non-injectivity} (v2; proved).
      In any tower $\Omega_0 \xrightarrow{\Pi_1} \Omega_1 \xrightarrow{\Pi_2} \Omega_2$ of
      genuinely distinct emergent levels, each $\Pi_i$ is non-injective with $S_{\Pi_i} > 0$,
      and $S_{\Pi_2 \circ \Pi_1} \geq S_{\Pi_1}$.
      Non-injectivity accumulates across levels: a framework with $k$ genuinely distinct
      levels carries at least $k-1$ independent non-injective projections.

      \textit{Structural colour confinement} (§6; structural).
      The tower $\Omega_0 \to \{\text{individual sectors } c_i\} \to \{\text{colour-neutral pairs and triplets}\} \to \{\text{leptons, hadrons}\}$
      instantiates Corollary~6 at the colour stratum.
      The colour projection $\Pi_2: \{c_i\} \to \{\text{colour-neutral}\}$ is necessarily
      non-injective.
      The non-observability of free colour charges is therefore a structural consequence
      of $S_{\Pi_2} > 0$, not a dynamical accident.

      \textit{Algebraic complementary route} (Remark~5; structural).
      In frameworks whose fibre carries a Heisenberg algebra $[X, Y] = Z \neq 0$, non-trivial
      kernel and $\SPI > 0$ follow directly from the algebraic structure.
      This is complementary to Theorem~3: the abstract no-go answers \emph{what} is forced;
      the algebraic route answers \emph{why} in frameworks with Heisenberg fibres.

    \subsection{Axiomatic foundation: Foundation~M}
      \label{sec:foundation}

      \textbf{Foundation~M}~\cite{Beau2026m} establishes the minimal axiomatic basis and
      derives from A1--A4 alone the following chain of results.

      The four axioms (§2):
      \begin{itemize}
        \item A1 (Local projective admissibility): there exists a non-empty set $F_n$ of admissible
        successor directions from any observable state $O_{n-1}$.
        \item A2 (Structural non-injectivity): distinct admissible directions may lead to the same
        resolved state; $\Pi_n$ is generically non-injective.
        \item A3 (Non-premature selection): an admissible transition preserves the full multiplicity
        of open directions until projection locking; no premature selection among them is allowed.
        \item A4 (Projection locking): resolution occurs only when continuous BI saturation meets
        the discrete shell support; the transition from admissible to admitted is structurally discrete.
      \end{itemize}

      Derived results (proved):
      \begin{itemize}
        \item Irreversibility and the arrow of time (Proposition~4.1 and Theorem~4.2): from A1+A2
        alone.
        \item Proto-state as physically real unresolved configuration (Proposition~5.2): from A3.
        \item Non-trivial commutator $[X, \sigma(X)] = Z \neq 0$ (Lemma~5.6 and its corollary):
        from A3 and irreducibility.
        \item Heisenberg group $\Heis_3(\mathbb{Z}/q\mathbb{Z})$ as symmetry group of $F_n$
        (Theorem~5.7): from A1--A3 and BI parity.
        \item $F_n \cong \Vr$ (Weil representation; stated in §5.4, detailed in HeisenbergStructure):
        from Theorem~5.7 and Stone--von Neumann.
        \item Discrete character of quantum transitions (Theorem~5.12): from A4 and O22.
        \item Projective incompleteness (Corollary to Theorem~5.4): every admissible $\Pi_n$ has
        non-trivial kernel; $[X, \sigma(X)] \neq 0$ structurally forbids trivial kernel.
      \end{itemize}

      Three roles of $\Pi_n$ (Remark~3.5; structural):
      admissibility filter, generator of temporal order, and revelation operator.
      These three roles are algebraically inseparable and all sourced in the non-commutativity.

    \subsection{Heisenberg structure from admissibility: HeisenbergStructure}
      \label{sec:heisstr}

      \textbf{HeisenbergStructure}~\cite{Beau2026HeisenbergStructure} provides the detailed proof
      of Foundation~M Theorem~5.7, omitted from Foundation for concision.
      The central result (Theorem~4.1; proved):
      \begin{quote}
        Under A1--A3 and Born--Infeld parity, the symmetry group of $F_n$ is isomorphic to
        $\Heis_3(\mathbb{Z}/q\mathbb{Z})$ for some prime $q$, and $F_n \simeq \Vr$.
      \end{quote}
      The proof rests on a single structural lemma (Lemma~3.1: admissible non-factorisability),
      followed by a chain of classification steps: non-trivial commutator, central commutator,
      Heisenberg algebra, prime order of the centre, Stone--von Neumann identification.

      As a corollary, the Heisenberg uncertainty principle is derived as a structural property
      of the unresolved proto-state, not as an independent postulate.

      The proof requires no input beyond A1--A3, BI parity, and classical group-classification
      results.
      It is entirely self-contained relative to Foundation~M and the BornInfeld paper.

    \subsection{No external dimensional scale: noscale}
      \label{sec:noscale}

      \textbf{noscale}~\cite{Beau2026j} establishes a structural rigidity result:
      the admissibility structure is closed under reparameterisations of $\cchi$ but not
      under the introduction of independent dimensional parameters.

      \textit{Main theorem} (Theorem~3.1; proved).
      Any deformation of the admissibility constraints that introduces an independent
      dimensional parameter $\lambda$ (with $[\lambda] \neq [\cchi]^k$ for any $k \in \mathbb{Q}$)
      breaks at least one of:
      (i) BFS-consistent spectral scaling of $\sigma_{\mathrm{pair}}(n)$;
      (ii) structural non-injectivity (A2) and the ENI no-go;
      (iii) bounded admissibility closure (A3--A4).

      \textit{Corollary} (proved).
      Dynamical fields carrying an independent dimensional coupling cannot be consistently
      coupled to the projection $\Pi$.
      In particular, a charge $Q$ with dimensions [charge] or a cosmological constant $\Lambda$
      with dimensions [length]$^{-2}$ cannot enter the admissibility structure.
      This provides the structural backing for the exclusion of Reissner--Nordstr\"{o}m and
      Schwarzschild--de Sitter geometries in Q6b.

      The Level~2 statement --- that every emergent scale is uniquely determined as a functional
      of $\cchi$ and the cascade --- is identified as an open problem (§5 of noscale).

      \begin{table}[ht]
        \centering
        \caption{Summary of constituent papers.
        Status: P = proved; S = structural.}
        \label{tab:papers}
        \renewcommand{\arraystretch}{1.2}
        \begin{tabular}{@{}lllll@{}}
          \toprule
          Paper & Central output & Status \\
          \midrule
          ENI & $\Pi$ non-injective $\Leftrightarrow$ $S_\Pi > 0$ $\Leftrightarrow$ genuine emergence & P \\
          ENI Cor.~6 & Recursive non-injectivity; colour confinement structural & P / S \\
          Foundation~M & A1--A4 $\Rightarrow$ $[X,Y]=Z\neq 0$ $\Rightarrow$ $\Heis_3$ $\Rightarrow$ $\Vr$ & P \\
          HeisenbergStructure & Detailed proof of Thm~5.7; uncertainty principle from A3 & P \\
          noscale & No independent dimensional parameter beyond $\cchi$ & P \\
          \bottomrule
        \end{tabular}
      \end{table}

  \section{Inputs from Upstream Results}
    \label{sec:inputs}

    The foundations sub-programme has no upstream dependencies within the corpus.
    It is the axiomatic root.

    The sole external input is the Born--Infeld uniqueness result of
    BornInfeld~\cite{Beau2026c}: the BI saturation functional is the unique bounded
    completion of the admissible effective action.
    This is established independently of A1--A4 and is taken as given by the
    foundations sub-programme when deriving the parity involution and the identification
    of $\Vr$.

  \section{Outputs to Downstream Branches}
    \label{sec:outputs}

    \begin{table}[ht]
      \centering
      \caption{Principal outputs and downstream consumers.}
      \label{tab:outputs}
      \renewcommand{\arraystretch}{1.3}
      \begin{tabular}{@{}p{5.8cm}p{4.2cm}l@{}}
        \toprule
        Output & Consumer & Status \\
        \midrule
        $\Pi$ non-injective; $\SPI > 0$ & Note~4 (Gravity), Q12, Q13 & P \\
        Arrow of time; irreversibility & Note~2 (Geometry), Q-series & P \\
        $[X,Y] = Z \neq 0$ (non-commutativity) & Q1, Q2, HeisStr & P \\
        $\Heis_3(\mathbb{Z}/q\mathbb{Z})$ as fibre group & All Notes 1--3 & P \\
        $\Vr \simeq L^2(\mathbb{Z}/q\mathbb{Z})$ (Weil repr.) & Notes 1--3; Q5a--Q11 & P \\
        BI bound $A_n \leq \cchi/\sqrt{\lambda_n}$ & Note~1, O-series & P \\
        Parity involution $c \leftrightarrow q-c$ & Note~1, O18, O30 & P \\
        No independent dimensional scale beyond $\cchi$ & Note~2 (noscale), Q6b & P \\
        Colour confinement as structural necessity & ENI Cor.~6, O31, O32 & S \\
        Projective incompleteness & Note~3 (gauge freedom) & P \\
        \bottomrule
      \end{tabular}
    \end{table}

  \section{Status of Results}
    \label{sec:status}

    \subsection{Proved results}
      \label{sec:proved}

      \begin{enumerate}
        \item \textit{ENI Theorem~3}: $\Pi$ non-injective $\Leftrightarrow$ genuine emergence,
        proved from surjectivity, informational completeness, non-isomorphism.
        Framework-independent.

        \item \textit{ENI Corollary~6}: recursive non-injectivity; any tower of $k$ distinct
        emergent levels carries at least $k-1$ independent non-injective projections.

        \item \textit{Irreversibility} (Foundation Proposition~4.1): from A1+A2; the arrow of time
        is not postulated.

        \item \textit{Temporal ordering} (Foundation Theorem~4.2): from irreversibility and novelty.

        \item \textit{Non-factorisability Lemma~5.6} (Foundation/HeisenbergStructure): A3 forces
        irreducibility of $F_n$; any stable decomposition prior to projection locking is forbidden.

        \item \textit{Non-trivial commutator} (Foundation Corollary to Thm~5.4):
        $[X, \sigma(X)] = Z \neq 0$ follows from irreducibility and the algebraic minimality of
        the admissible generators.

        \item \textit{$\Heis_3(\mathbb{Z}/q\mathbb{Z})$} (Foundation Theorem~5.7,
        HeisenbergStructure Theorem~4.1): the fibre symmetry group is uniquely the Heisenberg
        group of prime order; follows from the commutator structure and BI parity.

        \item \textit{$F_n \cong \Vr$} (HeisenbergStructure Theorem~4.1):
        Stone--von Neumann uniqueness of the irreducible representation with central character
        $e^{2\pi i/q}$.

        \item \textit{Uncertainty principle from A3} (HeisenbergStructure Corollary):
        $\Delta X \cdot \Delta Y \geq \hbar_{\mathrm{eff}}/2$ is a structural property of the
        unresolved proto-state; not an independent postulate.

        \item \textit{Discrete transitions} (Foundation Theorem~5.12): from A4 and O22.

        \item \textit{Projective incompleteness}: every admissible $\Pi_n$ has non-trivial kernel;
        no observable description is ever complete.

        \item \textit{No independent dimensional parameter} (noscale Theorem~3.1):
        proved for all three cases (global multiplicative, depth-dependent, configuration-dependent
        deformations).
      \end{enumerate}

    \subsection{Structural results}
      \label{sec:structural}

      \begin{enumerate}
        \item \textit{Structural colour confinement} (ENI §6):
        colour confinement as a projective necessity of the emergence tower.
        The quantitative dynamics of confinement (string tension, running coupling) is not
        derived here; only the structural necessity of non-observability is established.

        \item \textit{Three roles of $\Pi_n$} (Foundation Remark~3.5):
        admissibility filter, generator of temporal order, revelation operator.
        These roles are algebraically inseparable and all sourced in non-commutativity.

        \item \textit{Gauge freedom as projective redundancy}:
        gauge freedom is not a separately postulated symmetry principle; it is the residual
        fibre symmetry of the non-injective projection.
        This is structurally stated in Foundation and detailed in Note~3.

        \item \textit{Level~2 scale determination} (noscale §5; open):
        that every emergent scale is uniquely determined as a functional of $\cchi$ and the
        cascade is identified as an open structural problem, beyond the Level~1 rigidity proved
        in noscale.
      \end{enumerate}

    \subsection{Open hypotheses}
      \label{sec:open-hyp}

      \begin{enumerate}
        \item \textit{Born rule for general observables.}
        The Born rule is proved in the SU(2) sector for the co-admissible spin-$\tfrac{3}{2}$
        sector on $2I$~\cite{Beau2026q3}.
        Extension to arbitrary observables, multipartite systems, and continuous spectra
        requires extending the fibre structure beyond the two-element parity case.

        \item \textit{Full Hilbert structure from discrete fibre.}
        The passage from the discrete $\Vr \simeq L^2(\mathbb{Z}/q\mathbb{Z})$ to a continuous
        Hilbert space is handled by the Q5a continuum limit.
        The foundations sub-programme does not derive this passage; it delegates it to
        the Emergent Geometry sub-programme.

        \item \textit{Level~2 scale determination (noscale)}.
        That all emergent scales (masses, coupling constants, cosmological constant) are
        uniquely determined from $\cchi$ is open.
        Only Level~1 (no independent scale can be introduced without breaking admissibility)
        is proved.
      \end{enumerate}

  \section{Open Deliverables}
    \label{sec:open}

    Two open deliverables bound the sub-programme.

    \textbf{Open deliverable 1: Born rule for general observables.}
    The born rule is established within the SU(2) sector.
    Its extension to arbitrary observables and to multipartite systems at the level of the
    foundations axioms is the primary open structural problem of this sub-programme.

    \textbf{Open deliverable 2: Level~2 scale determination.}
    Proving that every emergent scale is uniquely determined from $\cchi$ and the cascade
    (noscale Level~2) would make the framework fully predictive with a single dimensional
    free parameter.
    This requires a structural argument that the present programme does not yet have.

  \section{Conclusion}
    \label{sec:conclusion}

    The non-injective foundations sub-programme derives the axiomatic primitive of the
    Cosmochrony corpus.
    Starting from four axioms (A1--A4) governing admissible transitions, it establishes
    as theorems: the structural necessity of non-injectivity (ENI); the arrow of time
    and irreversibility; the Heisenberg group $\Heis_3(\mathbb{Z}/q\mathbb{Z})$ and its
    Weil representation $\Vr$ as the unique admissible fibre; and the absence of any
    independent dimensional parameter beyond $\cchi$.

    The white-paper vocabulary of $\chi$, relaxation, and iterated projection is superseded.
    The substrate is static; the admissibility constraint replaces relaxation throughout the
    corpus.

    These results are the inputs consumed silently by every other sub-programme.
    Reading the corpus without this note is possible but the origin of each structural
    input remains opaque.
    This note makes that origin explicit: every structure that the other sub-programmes
    compute is a theorem of A1--A4, not an independent assumption.

    \newpage
    \bibliographystyle{plain}
    \bibliography{cosmochrony-bibliography}

\end{document}
